---
title: "Mainland ABC"
output: 
  github_document:
    toc: true
always_allow_html: true
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(rmarkdown)
library(usmap)
library(data.table)
library(scales)
library(janitor)
library(kableExtra)
```


## Samples 
```{r Samples, echo=FALSE, fig.height=3.5, fig.width=5.5, message=FALSE, warning=FALSE}

usa <- map_data("state")

east6emp <- c("SRR24465306","SRR24465269","SRR24465307","SRR24465297","SRR24465272","SRR24465305")

west6emp <- c("SRR24465296","SRR24465288","SRR24465292","SRR24465293","SRR24465285","SRR24465294")

east6dem <- c("SRR24465268","SRR24465309","SRR24465270","SRR24465308","SRR24465271","SRR24465304")

west6dem <- c("SRR24465287","SRR24465284","SRR24465283","SRR24465290","SRR24465289","SRR24465291")


pop <- read_tsv("inputstats_files/mainland_WGSmetadata.txt") %>% 
  dplyr::rename("long"=Lon,"lat"=Lat) %>% 
  filter(Dataset=="Sacks") %>% 
  filter(`Depth(x)`>4.5) %>% 
  filter(Deme=="pure")   %>% 
  mutate(Subset=case_when(SRA %in% east6emp ~ "Empirical",
                          SRA %in% west6emp ~ "Empirical",
                          TRUE ~ "Priors")) 



samps <- cbind(east6emp, west6emp, east6dem, west6dem) %>% unname()

kable(samps, "html", escape = FALSE) %>% 
  kable_styling(full_width = F) %>%
  add_header_above(c("East" = 1, "West" = 1,"East" = 1, "West" = 1)) %>% 
  add_header_above(c("Empirical" = 2, "Priors" = 2))

ggplot() +
  geom_polygon(data = usa, aes(x=long, y = lat, group = group), 
               fill = "white", color="gray70", linewidth=0.3) + 
  geom_point(data=pop, aes(x=long, y = lat,fill=Region, color=Subset),size=2, stroke=0.5,shape=21) +
  coord_fixed(xlim = c(-123, -75),  ylim = c(26, 41.8)) +
  scale_fill_manual(values = c("goldenrod2", "aquamarine3"))+
  scale_color_manual(values = c("black", "red"))+
  xlab("") + ylab("") +
  theme(panel.grid.minor=element_blank(),
        panel.grid.major = element_blank(),
        legend.position = c(0.15,0.15),
        legend.key = element_rect(colour = NA, fill = NA),
        legend.spacing.y = unit(-0.1, "cm"),
        legend.title = element_blank(),
        panel.background=element_rect(color="white",fill="white",linewidth = 0.5),
        axis.line=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank()) +
  guides(colour = guide_legend(nrow = 1),fill = guide_legend(nrow = 1))


```


## Priors

### SMC++ output
```{r Priors, echo=FALSE, fig.height=3,fig.width=5, message=FALSE, warning=FALSE}
ploteast <- read_csv("ABC_files/plot_east_mask.csv")

plotwest <- read_csv("ABC_files/plot_west_mask.csv")

pb <- bind_rows("east"=ploteast,"west"=plotwest,.id = "pop") %>% 
  filter(x>1e3 & x<1e6) %>% select(pop,x,y)

zx_breaks = c(1e3, 1e4, 1e5, 1e6)
zy_breaks = c(0,25e3,5e4,75e3,1e5,15e4,2e5,5e5)

ggplot()+
  geom_line(data=pb,aes(x=x,y=y,color=pop),size=1) + 
  scale_y_continuous(trans = "log",breaks=zy_breaks)+  
  scale_x_continuous(trans = "log",breaks=zx_breaks)+ 
  labs(x ="Generations", y = "Effective Population Size") +
  theme_classic() +
  theme(axis.text=element_text(color="black", size=11),
        legend.position = c(0.9,0.2)) +
  scale_color_manual(values = c("goldenrod2","aquamarine3")) 


eastroundpb <- pb %>% filter(pop=="east") %>% 
  dplyr::group_by(y) %>% 
  dplyr::rename("Ne"="y") %>% 
  dplyr::summarise(time_end=max(x),time_st=min(x)) %>%  
  arrange(desc(time_st)) %>%
  mutate(Ne=plyr::round_any(Ne,1000), 
         time_st=plyr::round_any(time_st,1000), 
         time_end=plyr::round_any(time_end,1000))


westroundpb <- pb %>% filter(pop=="west") %>% 
  dplyr::group_by(y) %>% 
  dplyr::rename("Ne"="y") %>% 
  dplyr::summarise(time_end=max(x),time_st=min(x)) %>%  
  arrange(desc(time_st)) %>%
  mutate(Ne=plyr::round_any(Ne,1000), 
         time_st=plyr::round_any(time_st,1000), 
         time_end=plyr::round_any(time_end,1000))

roundpb <- cbind(eastroundpb,westroundpb) %>% unname()


kable(roundpb, "html", escape = FALSE) %>% 
  kable_styling(full_width = F) %>%
  add_header_above(c("Ne" = 1, "end" = 1,"start" = 1,"Ne" = 1, "end" = 1,"start" = 1)) %>% 
  add_header_above(c("East" = 3, "West" = 3))
```



### Distributions for Sampling
```{r DemParam, echo=FALSE, fig.height=4, message=FALSE, warning=FALSE}
priors <- read_tsv("ABC_files/priors_t16.txt") 


kable(priors, "html", escape = FALSE,col.names = NULL) %>% 
  kable_styling(full_width = F) %>%
  add_header_above(c("Variable" = 1, "Min" = 1,"Max" = 1,"Min" = 1,"Max" = 1)) %>%
  add_header_above(c(" "=1, "East" = 2, "West" = 2))

```

### Recombination Rates

Obtained from running pyrho (and SMC++) on east (n=14) and west (n=12) separately

Genome-wide recombination rate average and standard deviation used to determine gamma distribution for sampling

**East**

> RecRateBp_east = random.gammavariate(0.5591, 0.00000002403)


**West**

> RecRateBp_west = random.gammavariate(0.9469, 0.00000002202)


## Neutral Genome

From dog annotation liftover, identify canoncial coding sequence coordinates to be removed


### Genic Regions

create bed file of genes ±1kb from annotation

<details>
<summary>Show code</summary>
<br>
```{r annotation, eval=FALSE, warning=FALSE,message=FALSE}

#Grayfox liftover annotation

galba <- fread("~/Downloads/galba.gtf.gz")



chroms <- read_tsv("inputstats_files/grayfox_renameChroms_number.txt", col_names = c("scaf","chrom")) %>% 
  separate(scaf, remove=F, c(NA,NA,NA,NA,NA,"length"))

gfgenes <- galba %>% left_join(chroms,by=c("V1"="scaf"))


mybednew <- gfgenes %>% 
  filter(str_detect(V3, "CDS")) %>% mutate(len=V5-V4) %>% separate(V9, sep=";", into=c("t","g"))

mybeds <- mybednew %>%  mutate(g = str_remove_all(g, "\"")) %>% separate(g, sep="d", into=c("id","gid")) %>% mutate(gid=as.numeric(str_remove(gid,"g"))) %>% arrange(gid,len)

mb <- mybeds %>% group_by(gid) %>% filter(len==max(len)) %>% drop_na(chrom) %>% ungroup()


mb1kb %>% write_tsv("sumstats_files/gfgenes_2024_1kb.bed", col_names = F)

## take the intersect to get neutral regions


neutral <- read_tsv("sumstats_files/neutral.bed", col_names = c("chrom","start","end")) %>% mutate(size=end-start) 

neutral %>% filter(size>1000)  %>% dplyr::select(chrom,start,end) %>%  write_tsv("sumstats_files/neutral_regions_gt1kb.bed", col_names = F)

neutralsmall <- neutral %>% filter(size<1000) %>% dplyr::select(chrom,start,end)

mbgt1kb <- bind_rows(mb1kb,neutralsmall) %>% arrange(chrom,start)

mbgt1kb %>% write_tsv("sumstats_files/gfgenes_2024_1kb_gt1kb.bed", col_names = F)

```


</details>



```{r plotNeutral, echo=FALSE, fig.height=4, fig.width=7, message=FALSE, warning=FALSE}

neutral_reg <- read_tsv("ABC_files/neutral_regions_gt1kb.txt",col_names = "size") %>% 
  mutate(cutsize=cut(size,c(1e3,5e3,1e4,5e4,1e5,5e5,1e6,2e6,Inf),
                     labels=c('1kb-5kb','5kb-10kb','10kb-50kb','50kb-100kb','100kb-500kb','500kb-1mb','1mb-2mb','>2mb')))


neutral_tab <- neutral_reg %>% group_by(cutsize) %>% count() 



ggplot(neutral_tab)+
  geom_col(aes(x=cutsize,y=n))+
  theme_classic()+
  theme(
    panel.grid.minor=element_blank(),
    panel.grid.major=element_blank(),
    panel.background=element_blank(),
    strip.background = element_blank(),
    axis.text = element_text(colour = "black", size = 10),
    axis.line = element_line(colour = "black", linewidth = 0.5),
    axis.ticks = element_line(colour = "black", linewidth = 0.5),
    axis.title = element_blank())


```



## Summary Statistics


Do for empirical subset of samples (n=6) per pop

```{r empBin, echo=FALSE, fig.height=4, fig.width=5, message=FALSE, warning=FALSE}

empbin<-fread("ABC_files/emp_bin_summary.txt")


ggplot(empbin)+
  geom_col(aes(x=bin,y=count,fill=pop))+
  facet_wrap(~stat+pop,scales="free")+
  scale_fill_manual(values = c("goldenrod2","aquamarine3")) +
  theme_classic()+
  theme(
    panel.grid.minor=element_blank(),
    panel.grid.major=element_blank(),
    panel.background=element_blank(),
    strip.background = element_blank(),
    axis.text = element_text(colour = "black", size = 10),
    axis.line = element_line(colour = "black", linewidth = 0.5),
    axis.ticks = element_line(colour = "black", linewidth = 0.5),
    axis.title = element_blank(),
    legend.position = "none")+  
  scale_y_continuous(labels = ~ format(.x, scientific = FALSE))+
  scale_x_continuous(breaks = seq(1, 6, by = 1))
```




#### Generate site frequency spectra for east and west populations

<details>
<summary>Show code for getting allele counts</summary>
<br>


```{bash alleles, eval=F}

##Filter for six empirical samples per pop

bcftools view -Oz -S west6.txt /project/jazlynmo_738/DataRepository/Canids/Variants/GrayFox/Mainland/grayfox_filtered.renameChroms.Mainland.drop295.ACgr59_DPgr205lt500.vcf.gz > west6.vcf.gz

bcftools view -Oz -S east6.txt /project/jazlynmo_738/DataRepository/Canids/Variants/GrayFox/Mainland/grayfox_filtered.renameChroms.Mainland.drop295.ACgr59_DPgr205lt500.vcf.gz > east6.vcf.gz

bcftools index -t west6.vcf.gz

bcftools index -t east6.vcf.gz

##Filter for SNPs among sample set
bcftools view --min-ac=1 --max-ac 11 west6.vcf.gz | bcftools view -i 'F_MISSING<0.1' -Oz -o snpwest6.vcf.gz

bcftools index -t snpwest6.vcf.gz

bcftools view --min-ac=1 --max-ac 11 east6.vcf.gz | bcftools view -i 'F_MISSING<0.1' -Oz -o snpeast6.vcf.gz

bcftools index -t snpeast6.vcf.gz


##Get allele counts
vcftools --gzvcf snpeast6.vcf.gz --counts --out eastcount_neutral --exclude-bed gfgenes_2024_1kb_gt1kb.bed >& eastcount_neutral.log

vcftools --gzvcf snpwest6.vcf.gz --counts --out westcount_neutral --exclude-bed gfgenes_2024_1kb_gt1kb.bed >& westcount_neutral.log

sed -e 's/:/\t/g'  eastcount_neutral.frq.count > east_neutral.count

sed -e 's/:/\t/g'  westcount_neutral.frq.count > west_neutral.count




```

</details>

<br>



#### Bin allele counts to get site frequency spectrum

<details>
<summary>Show code for binning allele counts</summary>
<br>


```{r binAlleles, eval=F}

eastcount <- fread("/scratch1/marjanak/mainland_2024/gf_correct/east_neutral.count",col.names = c("CHROM","POS","N_ALLELES","N_CHR","allele1","count1","allele2","count2"))

sumeast <- eastcount %>% group_by(count1,count2) %>% count()

east_data <- sumeast %>%
  group_by(bin = pmin(count1, count2), bin2 = pmax(count1, count2)) %>%
  summarise(count = sum(n), .groups = 'drop') %>%
  arrange(bin) %>% select(bin,count)

westcount <- fread("/scratch1/marjanak/mainland_2024/gf_correct/west_neutral.count",col.names = c("CHROM","POS","N_ALLELES","N_CHR","allele1","count1","allele2","count2"))

sumwest <- westcount %>% group_by(count1,count2) %>% count()

west_data <- sumwest %>%
  group_by(bin = pmin(count1, count2), bin2 = pmax(count1, count2)) %>%
  summarise(count = sum(n), .groups = 'drop') %>%
  arrange(bin) %>% select(bin,count)

snpbin <- bind_rows("east"=east_data, "west"=west_data, .id="pop")

statsbin <- bind_rows("pi"=pibin,"sfs"=snpbin,.id="stat")

write_tsv(statsbin,"/scratch1/marjanak/mainland_2024/gf_correct/emp_bin_summary.txt")


```

</details>

<br>



#### Calculate per-site nucleotide diversity (pi) in east and west using gVCFs 


<details>
<summary>Show code for per chrom job array</summary>
<br>

```{bash sitePi, eval=F}
#!/bin/sh
#SBATCH --job-name=sitepiE
#SBATCH --output=/scratch1/marjanak/mainland_2024/gf_correct/pi_east/slurmJob_%a.out
#SBATCH --error=/scratch1/marjanak/mainland_2024/gf_correct/pi_east/slurmJob_%a.err
#SBATCH --time=30:00:00
#SBATCH --partition=qcb
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=4
#SBATCH --mem-per-cpu=8000MB
#SBATCH --array=1-32
#SBATCH --mail-type=END,FAIL # notifications for job done & fail
#SBATCH --mail-user=marjanak@usc.edu

module load vcftools

vcftools --gzvcf /project/jazlynmo_738/DataRepository/Canids/Invariant/GrayFox/Mainland/grayfox_filtered.renameChroms.Mainland.drop295.ACgr59_DPgr205lt500.rmMultAllelicgvcf.gz --keep east6.txt --site-pi --max-missing 1.0 --chr chr${SLURM_ARRAY_TASK_ID} --out /scratch1/marjanak/mainland_2024/gf_correct/pi_east/chr${SLURM_ARRAY_TASK_ID}sitepi_East

sleep 60
```

</details>

<br>

#### Calculate nucleotide diversity in neutral regions

<details>
<summary>Show code for getting nucelotide diversity</summary>
<br>
```{bash sumPi, eval=F}

#Skip header and combine site pi per chrom outputs
awk 'FNR>1' chr* > east_sites.pi

#Delete lines with pi of zero
awk '$3 != 0' east_sites.pi > allChroms.sites.pi

#Convert to bed format
awk 'OFS="\t"{print $1, $2-1, $2, $3}' allChroms.sites.pi > allChroms.sites.bed


#Intersect per site pi and neutral region bed files to calculate avg pi per region 
bedtools intersect -a neutral_regions_gt1kb.bed -b allChroms.sites.bed -wa -wb > intersected_sites.bed
```
</details>

<br>




#### Bin Nucelotide Diversity


<details>
<summary>Show code for binning nucleotide diversity</summary>
<br>
```{r pibin, eval=F}

eastint <- read_tsv("/scratch1/marjanak/mainland_2024/gf_correct/pi_east/intersected_sites.bed",
                    col_names = c("chr","start","end","chrom","x","y","pi"))

westint <- read_tsv("/scratch1/marjanak/mainland_2024/gf_correct/pi_west/intersected_sites.bed",
                    col_names = c("chr","start","end","chrom","x","y","pi"))


allneut <- read_tsv("/scratch1/marjanak/mainland_2024/gf_correct/pi_east/neutral_regions_gt1kb.bed",
                    col_names = c("chr","start","end"))

eastneut <- allneut %>% 
  left_join(eastint, by = c("chr","start","end")) 

counteast <- eastneut %>% 
  mutate(pie=case_when(is.na(pi) ~ 0, TRUE ~ pi)) %>% 
  group_by(chr,start,end) %>% 
  summarise(n=n(), sumpi=sum(pie)) %>% 
  mutate(size=end-start, avgpi=sumpi/size) %>% 
  mutate(cutsize=cut(avgpi,c(-Inf,0.0008,0.0016,0.0024,0.0032,0.004,Inf),
                     labels = c("nd_1","nd_2","nd_3","nd_4","nd_5","nd_6")))

cte <- counteast %>% group_by(cutsize) %>% summarise(regs=n())

westneut <- allneut %>% 
  left_join(westint, by = c("chr","start","end")) 

countwest <- westneut %>% 
  mutate(pie=case_when(is.na(pi) ~ 0, TRUE ~ pi)) %>% 
  group_by(chr,start,end) %>% 
  summarise(n=n(), sumpi=sum(pie)) %>% 
  mutate(size=end-start, avgpi=sumpi/size) %>% 
  mutate(cutsize=cut(avgpi,c(-Inf,0.0017,0.0034,0.0051,0.0068,0.0085,Inf), 
                     labels = c("nd_1","nd_2","nd_3","nd_4","nd_5","nd_6")))

ctw <- countwest %>% group_by(cutsize) %>% summarise(regs=n())

pibin <- bind_rows("east"=cte, "west"=ctw, .id="pop") %>% mutate(bin=as.integer(bin))

```

</details>

<br>