---
title: "input"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(rmarkdown)
library(usmap)
library(data.table)
```

### Samples


focus on mainland foxes from Sacks only (remove one bad sample) *n=41 remaining*

figure out number of non-hybrid samples  *n=26 (14 east + 12 west)*

```{r Samples, warning=FALSE, message=FALSE, echo=FALSE, fig.height=3, fig.width = 7}

usa <- map_data("state")

pop <- read_tsv("inputstats_files/mainland_WGSmetadata.txt") %>% 
  dplyr::rename("long"=Lon,"lat"=Lat) %>% 
  mutate(`Samples (n=41)`=case_when(Deme=="hybrid" ~ "hybrid (15)",
                                    Region=="West" ~ "west (12)",
                                    TRUE ~ "east (14)")) %>% 
  mutate(depth=case_when(`Depth(x)`>18 ~ 8,
                         TRUE ~ 4)) %>% 
  filter(Dataset=="Sacks")


ggplot() +
  geom_polygon(data = usa, aes(x=long, y = lat, group = group), 
               fill = "gray90", color="gray40", linewidth=0.3) + 
  geom_point(data = pop, aes(x=long, y = lat, fill = `Samples (n=41)`),shape = 21, size=2) +
  coord_fixed(xlim = c(-123, -75),  ylim = c(26, 41.8)) +
  scale_fill_manual(values = c("dodgerblue2", "goldenrod", "red2"))+
  xlab("") + ylab("") +
  theme(panel.grid.minor=element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "top",
        legend.title = element_text(size=9),
        panel.background=element_rect(color="white",fill="white",linewidth = 0.5),
        axis.line=element_line(color="black"),
        axis.text=element_text(color="black"),
        axis.ticks=element_line(color="black")) 


pop  %>% filter(Deme=="pure") %>%  group_by(`Samples (n=41)`, Sex) %>% count()
```

split data in half 

```{r subSamples, warning=FALSE, message=FALSE, echo=FALSE, fig.height=3, fig.width = 7}

ggplot() +
  geom_polygon(data = usa, aes(x=long, y = lat, group = group), 
               fill = "gray90", color="gray40", linewidth=0.3) + 
  geom_point(data = subset(pop, Deme=="pure"),  
             aes(x=long, y = lat, shape=Sex, size=`Depth(x)`), fill="black") +
  coord_fixed(xlim = c(-123, -75),  ylim = c(26, 41.8)) +
  scale_shape_manual(values=c(21,24))+
  xlab("") + ylab("") +
  theme(panel.grid.minor=element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "top",
        legend.title = element_text(size=9),
        panel.background=element_rect(color="white",fill="white",linewidth = 0.5),
        axis.line=element_line(color="black"),
        axis.text=element_text(color="black"),
        axis.ticks=element_line(color="black")) 

#choose samples and create list

```


Samples for smc++ (higher depth)

|Region |SRA	        |Sex|State|	Lat.  |  Long.  | Depth |
| ----- | ----------- | - | --- | ----- | ------- | ----- | 
|	East  |	SRR24465304	|	M	|	SC	| 33.346|	-81.743	|	26.2  | 
| East	| SRR24465268	|	F	|	AR	| 35.708|	-92.519	|	7.2   |
|	East	| SRR24465307	|	M	|	AR	| 34.287|	-91.338	|	6.4   |
|	East	| SRR24465305	|	M	|	KY	| 37.322|	-84.928	|	6.3   |
|	East	| SRR24465269	|	M	|	AR	| 33.467|	-92.169	|	6.2   |
|	East	| SRR24465309	|	F	|	MO	| 37.152|	-91.391	|	6.2   |
|	West	| SRR24465290	|	M	|	TX	| 29.508|	-103.328|	23.6  |
|	West	| SRR24465296	|	M	|	CA	| 38.490|	-122.152|	7.7   |
|	West	| SRR24465294	|	M	|	TX	| 29.924|	-101.992|	6.4   |
|	West	| SRR24465289	|	M	|	TX	| 30.213|	-102.430|	6.1   |
|	West	| SRR24465291	|	F	|	TX	| 30.684|	-101.308|	6.1   |
|	West	| SRR24465287	|	M	|	NV	| 38.821|	-116.494|	5.8   |


Samples for empirical analysis

|Region |SRA	        |Sex|State|	Lat.  |  Long.  | Depth |
| ----- | ----------- | - | --- | ----- | ------- | ----- | 
| East  |	SRR24465272 | M | TN | 35.543 | -88.856 |  5.7  | 
| East  |	SRR24465270 | F | AR | 36.052 | -90.973 |  5.6  | 
| East  |	SRR24465308 | F | MS | 33.937 | -89.337 |  5.1  | 
| East  |	SRR24465306 | F | AR | 33.588 | -93.391 |  4.9  | 
| East  |	SRR24465297 | F | MS | 32.766 | -90.388 |  4.8  | 
| East  |	SRR24465271 | M | AL | 32.248 | -87.791 |  4.8  | 
| West  |	SRR24465292 | M | TX | 29.561 | -104.372|  5.5  | 
| West  |	SRR24465283 | F | NM | 33.156 | -104.321|  5.5  | 
| West  |	SRR24465293 | M | TX | 30.625 | -104.067|  5.5  | 
| West  |	SRR24465288 | F | NV | 37.459 | -114.460|  5.0  | 
| West  |	SRR24465285 | M | NM | 34.956 | -103.157|  5.0  | 
| West  |	SRR24465284 | M | NM | 32.711 | -107.845|  4.9  | 


### Genic Regions

create bed file of genes ±1kb from annotation

```{r annotation, eval=FALSE, warning=FALSE,message=FALSE}

# annotation file
galba <- fread("~/Downloads/galba.gtf.gz")

# chromosome name conversion file 
chroms <- read_tsv("grayfox_renameChroms_number.txt", col_names = c("scaf","chrom")) %>% 
  separate(scaf, remove=F, c(NA,NA,NA,NA,NA,"length"))

# write file with chromosome sizes
chroms %>% select(scaf,length) %>% write_tsv("chromSizes.txt", col_names = F)

# filter annotation for genic regions
genes <- galba %>% filter(V3=="gene") %>% select(V1,V4,V5)

# add chromosome names to annotation file and remove stuff not mapping to scaffs 1-32
genescaf <- genes %>% left_join(chroms,by=c("V1"="scaf")) %>% 
  select(chrom,V4,V5, length)  %>%  na.omit()

# add 1kb to either end of each gene, 
## make sure values don't drop below zero or exceed the length of the chromosome
gene1kb <- genescaf %>% mutate(start=V4-1000, end=V5+1000) %>% 
  mutate(newstart=case_when(start<0 ~ 1, TRUE ~ start)) %>% 
  mutate(newend=case_when(end>as.numeric(length) ~ as.numeric(length), TRUE ~ end))

gene1kb %>% select(chrom,newstart,newend) %>% write_tsv("genes1kb.bed", col_names = F)

```



remove genic regions and compute summary statistics on half of data



```{bash genic, eval=FALSE}

module load bedtools2
module load bcftools

bedtools sort -i genes1kb.bed > genicregions.1kb.bed

#removes genic regions
bcftools view -T ^genicregions.1kb.bed grayfox_filtered.renameChroms.Mainland.ACgr61_DPgr165lt500.vcf.gz -Oz -o grayfox_mainland_nogenes.vcf.gz 

bcftools index -t grayfox_mainland_nogenes.vcf.gz

#split by chromosome
bcftools index -s grayfox_mainland_nogenes.vcf.gz | cut -f 1 | while read C; do bcftools view -O z -o split.${C}.vcf.gz grayfox_mainland_nogenes.vcf.gz "${C}" ; done

```

### Compute Summary Statistics

calculate pi, segregating sites, heterozygosity per site on n=12 empirical samples


```{bash vcftools-stats, eval=FALSE}

#!/bin/sh
#SBATCH --job-name=statsvcfsite
#SBATCH --output=/scratch1/marjanak/statsvcfsite.out
#SBATCH --error=/scratch1/marjanak/statsvcfsite.err
#SBATCH --partition=qcb
#SBATCH --time=30:00:00
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=4
#SBATCH --mem-per-cpu=8000MB
#SBATCH --mail-type=END,FAIL # notifications for job done & fail
#SBATCH --mail-user=marjanak@usc.edu

module load vcftools

vcftools --gzvcf /project/jazlynmo_738/DataRepository/Canids/Invariant/GrayFox/Mainland/grayfox_filtered.renameChroms.Mainland.ACgr61_DPgr165lt500.gvcf.gz --keep east6.txt --exclude-bed genicregions.1kb.bed --site-pi --out east6_pi_site.out

vcftools --gzvcf /project/jazlynmo_738/DataRepository/Canids/Invariant/GrayFox/Mainland/grayfox_filtered.renameChroms.Mainland.ACgr61_DPgr165lt500.gvcf.gz --keep west6.txt --exclude-bed genicregions.1kb.bed --site-pi --out west6_pi_site.out


vcftools --gzvcf /project/jazlynmo_738/Maria/grayfox_mainland_nogenes.vcf.gz --keep east6.txt --snpdensity 10 --out east6_S_10bp.out

vcftools --gzvcf /project/jazlynmo_738/Maria/grayfox_mainland_nogenes.vcf.gz --keep west6.txt --snpdensity 10 --out west6_S_10bp.out



```

```{bash scratch, echo=FALSE, eval=FALSE}


#!/bin/sh
#SBATCH --job-name=sitesvcf
#SBATCH --output=/scratch1/marjanak/sitesvcf.out
#SBATCH --error=/scratch1/marjanak/sitesvcf.err
#SBATCH --partition=qcb
#SBATCH --time=24:00:00
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=4
#SBATCH --mem-per-cpu=8000MB
#SBATCH --mail-type=END,FAIL # notifications for job done & fail
#SBATCH --mail-user=marjanak@usc.edu

module load vcftools









vcftools --gzvcf /project/jazlynmo_738/DataRepository/Canids/Invariant/GrayFox/Mainland/grayfox_filtered.renameChroms.Mainland.ACgr61_DPgr165lt500.gvcf.gz --keep east6.txt --exclude-bed genicregions.1kb.bed --TajimaD 10000 --out east6_tajD_10kb.out

vcftools --gzvcf /project/jazlynmo_738/DataRepository/Canids/Invariant/GrayFox/Mainland/grayfox_filtered.renameChroms.Mainland.ACgr61_DPgr165lt500.gvcf.gz --keep west6.txt --exclude-bed genicregions.1kb.bed --TajimaD 10000 --out west6_tajD_10kb.out
```


Average nucleotide diversity 1.8x higher in the east (n=6) compared to the west (n=6)

```{r pi, warning=FALSE, message=FALSE, echo=FALSE, fig.height=4, fig.width = 10}

eastpi <- read_tsv("east6_pi_10kb.out.windowed.pi") %>%
  mutate(chr_position = ((BIN_START + BIN_END)/2)/1000000) %>%
  mutate(CHROM=as.numeric(str_remove(CHROM, "chr"))) 


westpi <- read_tsv("west6_pi_10kb.out.windowed.pi") %>%
  mutate(chr_position = ((BIN_START + BIN_END)/2)/1000000) %>%
  mutate(CHROM=as.numeric(str_remove(CHROM, "chr"))) 


allpi <- bind_rows("east"=eastpi,"west"=westpi,.id = "pop")

allpi %>% group_by(pop) %>% summarise(avgpi=mean(PI, notation="sci"))


ggplot() + 
  geom_line(data=allpi, aes(x=chr_position, y=PI, color=pop), size=0.15) +
  scale_alpha_manual(values = rep(c(1, 0.7), 100)) +
  scale_color_manual(values = c("#045acb", "#cf1a10"))+
  xlab("Position (Mbp)") +
  facet_grid(cols = vars(CHROM), rows=vars(pop),
             space = "free_x",
             scales = "free_x") +     
  ylab(expression(pi))+
  #scale_y_continuous(limits = c(0,0.025), breaks = c(0, 0.01, 0.02))+
  theme(
    panel.grid.minor=element_blank(),
    panel.grid.major=element_blank(),
    panel.background=element_blank(),
    axis.text = element_text(colour = "black", size = 12),
    axis.line.y = element_line(colour = "black", linewidth = 0.5),
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.ticks.y = element_line(colour = "black", size = 0.5),
    axis.ticks.length = unit(0.2, "cm"),
    axis.title.x = element_text(colour = "black", size = 14),
    legend.position="none",
    strip.text = element_text(colour = "black", size = 12),
    axis.title.y = element_text(colour = "black", size = 14),   
    panel.spacing = unit(0.0, "cm"),
    strip.background=element_rect(fill="gray90"),
    axis.text.x = element_blank()) + 
  geom_hline(yintercept=0.000108,color = "turquoise", size=.5) + 
  geom_hline(yintercept=0.0000611,color = "gold", size=.5)
  





cfeastpi <- read_tsv("cf_east6_pi_10kb.out.windowed.pi") %>%
  mutate(chr_position = ((BIN_START + BIN_END)/2)/1000000) %>%
  mutate(CHROM=as.numeric(str_remove(CHROM, "chr"))) 


cfwestpi <- read_tsv("cf_west6_pi_10kb.out.windowed.pi") %>%
  mutate(chr_position = ((BIN_START + BIN_END)/2)/1000000) %>%
  mutate(CHROM=as.numeric(str_remove(CHROM, "chr"))) 


cfallpi <- bind_rows("east"=cfeastpi,"west"=cfwestpi,.id = "pop")

cfallpi %>% group_by(pop) %>% summarise(avgpi=mean(PI, notation="sci"))


ggplot() + 
  geom_line(data=cfallpi, aes(x=chr_position, y=PI, color=pop), size=0.15) +
  scale_alpha_manual(values = rep(c(1, 0.7), 100)) +
  scale_color_manual(values = c("#045acb", "#cf1a10"))+
  xlab("Position (Mbp)") +
  facet_grid(cols = vars(CHROM), rows=vars(pop),
             space = "free_x",
             scales = "free_x") +     
  ylab(expression(pi))+
  #scale_y_continuous(limits = c(0,0.025), breaks = c(0, 0.01, 0.02))+
  theme(
    panel.grid.minor=element_blank(),
    panel.grid.major=element_blank(),
    panel.background=element_blank(),
    axis.text = element_text(colour = "black", size = 12),
    axis.line.y = element_line(colour = "black", linewidth = 0.5),
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.ticks.y = element_line(colour = "black", size = 0.5),
    axis.ticks.length = unit(0.2, "cm"),
    axis.title.x = element_text(colour = "black", size = 14),
    legend.position="none",
    strip.text = element_text(colour = "black", size = 12),
    axis.title.y = element_text(colour = "black", size = 14),   
    panel.spacing = unit(0.0, "cm"),
    strip.background=element_rect(fill="gray90"),
    axis.text.x = element_blank()) + 
  geom_hline(yintercept=0.0000800,color = "turquoise", size=.5) + 
  geom_hline(yintercept=0.000173,color = "gold", size=.5)
  


```





